---
title: "Baby Names"
author: "Giuseppe A. Paleologo"
date: "Tuesday, March 17, 2015"
output: pdf_document
---


## Read Files
```{r message=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(data.table)
library(stringr)
library(magrittr)
setwd('C:/Users/Giuseppe/Dropbox (Personal)/babynames')

years <- 1880:2011
path <- sprintf('names/yob%d.txt', years)
columns <- c('name', 'sex', 'births')

reader <- function(yr){
  x<- fread(sprintf('names/yob%d.txt', yr), data.table=FALSE) %>% set_names(columns)
  x$year <- yr
  x
}

babynames <- lapply(years, reader) %>%  bind_rows 
```

## Plots births by sex and year
```{r}
library(magrittr)
babynames %>%
  group_by(year, sex) %>% 
  summarize(births=sum(births)) %>%
  ggplot(aes(x=year, y=births, color=sex)) + geom_line()
````

## Fraction of total names, within sex and year
```{r}
babynames %<>% 
  group_by(year, sex) %>% 
  mutate(prop=births/sum(births)) 
````

## Top 1000 names by year, sex
```{r}
top1000 <- babynames %>% 
  group_by(year, sex) %>% 
  arrange(desc(births)) %>%
  filter(min_rank(desc(births)) <= 1000) 
top1000
```

## Plots the number of babies named John, Harry, Mary, Marilyn over time
```{r}
babynames %>% 
  filter(name %in% c('John', 'Harry', 'Mary', 'Marilyn')) %>%
  group_by(year, name) %>%
  summarize(births=sum(births)) %>%
  ggplot(aes(x=year, y=births)) + geom_line() + facet_grid(name ~ .)
```

## Plots the proportion of the top 1000 names as a percentage of total
```{r}
top1000 %>% 
  group_by(year, sex) %>% 
  summarize(prop=sum(prop)) %>%
  ggplot(aes(x=year, y=prop, color=sex)) + 
    geom_line() + 
    ggtitle('Sum of table1000.prop by year and sex') + 
    scale_y_continuous(limits=c(0, 1))
```

## How many boy names comprise 50% of the total in 2010?
```{r}
top1000 %>% 
  filter(year == 2010, sex =='M') %>% 
  arrange(desc(births)) %>%
  mutate(totprop = cumsum(prop)) %>%
  filter(totprop <= .50) %>%
  nrow
```

## Plots number of most popular names used by 50% of boys and girls over time
```{r}
get_quantile_count <- function(x, qtle=0.5) 
  x %>% sort(decreasing=TRUE) %>% cumsum %>% {.<= qtle} %>% sum

top1000 %>%
  group_by(year, sex) %>%
  summarize(No.names = get_quantile_count(prop)) %>%
  ggplot(aes(x=year, y=No.names, color=sex)) + 
    geom_line() + 
    ggtitle('Number of popular names in top 50%')
```

## Plot the distribution of names by last letter for three time snapshots
```{r}
letter_count <- babynames %>%
  mutate(last_letter = str_sub(name, start=-1L, end=-1L )) %>%
  group_by(year, sex, last_letter) %>% 
  summarise(count=sum(births))

letter_prop <- letter_count %>% 
  filter(year %in% c(1910, 1960, 2010)) %>%
  group_by(year, sex) %>%
  mutate(letter_prop=count/sum(count))

letter_prop %>% 
  ggplot(aes(x=last_letter, y=letter_prop, fill=as.factor(year))) + 
    geom_bar(stat='identity',position=position_dodge()) + 
    facet_grid(sex ~ .)
```

## Plots the proportion of boy names ending in 'd', 'n', and 'y' over time
```{r}
letter_count %>%
  filter(sex=='M') %>%
  group_by(year) %>%
  mutate(letter_count=sum(count), letter_prop=count/sum(count)) %>%
  filter(last_letter %in% c('d', 'n', 'y')) %>%
  ggplot(aes(x=year, y=letter_prop, color=last_letter)) +
    geom_line()   
```  
  
